---
title: "Comparing dominant height approaches"
author: "Brian Clough"
date: "08/25/2021"
output: html_document
editor_options:
  chunk_output_type: console
---

Lately I've been thinking a lot about how in forestry, many of the "measurements" we work with are really estimates. This came 
up in a recent project, where one of the goals was to estimate dominant height for a set of stands. We were seeing some discrepancies
among a couple different workups and, being a pretty experienced group of biometricians, we immediately asked each other how we'd 
defined dominant height when we did our calculations. Many of us know the definition of dominant height as "the average height of 
dominant and co-dominant trees". And if you've spent time working with inventory data, you know there's about a dozen different 
ways you can define it. Once we talked things through, we quickly got on the same page and reconciled the differences. 

This is a pretty common story in our field, and it comes up in a lot of different contexts (estimating volume, value, carbon, etc.).
It highlights the importance of an often underappreciated concept in statistics: the observation model. Simply put, the observation 
model links the thing you actually measure (trees on inventory plots) to value you're trying to approximate (dominant height of a stand).
In our example, the observation model includes the method we used to define dominant and co-dominant trees for the purpose of estimating 
plot, and in turn stand, dominant height. The choices you make when defining your observation model can have downstream consequences 
for your estimates and inferences. 

To illustrate, we'll expand on our original example of dominant height and look at how the choice of observation model impacts the 
development of site index curves. We'll work with FIA data, and for simplicity's sake we'll confine our comparison to Douglas-fir in 
Oregon. We'll consider two different observation models of site index, that will differ in the method used to estimate dominant height:
1. Define dominant height as the mean of all dominant and co-dominant trees as defined by FIA's crown class code (CCLD).
2. Define dominant height as the mean of all trees above the 80th percent quantile of each plot. 


We'll define age of the plot using FIA measurements of breast-height age (BHAGE). To understand the impact of the choice of 
observation model, we'll then fit some nonlinear height/age models to the data, using each of the formulations of dominant height 
as the response variable. If you want to step through the example, code can be in the forestry source Github repo.

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
remotes::install_github("SilviaTerra/tidyFIA")
library(tidyFIA)
library(brms)
library(tidybayes)
install.packages('tigris')
library(tigris)

divergent_palette <- rep(
  c(
    "#439E45", "#072649", "#DBE62F", "#6C7A8B", "#1CB3B5",
    "#E5330F", "#251EB1", "#CACEC6", "#93C7EE", "#C947D1",
    "#79FFE7", "#63340A"
  ),
  6
)

scale_colour_discrete <- function(...) {
  scale_colour_manual(..., values = divergent_palette)
}
scale_fill_discrete <- function(..., values = divergent_palette) {
  scale_fill_manual(..., values = values)
}

scale_fill_continuous <- ggplot2::scale_fill_viridis_c
scale_colour_continuous <- ggplot2::scale_colour_viridis_c

```

```{r query_and_filter}

or_fia <- tidy_fia(states = "OR", postgis = TRUE)

or_cond <- or_fia[["cond"]] %>% 
           filter(condprop_unadj > 0.99) %>% #Only single condition plots
           select(plt_cn, fortypcd)

df_plots <- or_fia[["plot"]] %>% 
            filter(plot_status_cd == 1) %>% #Only measured, forested plots 
            filter(invyr > 2016) %>% #Only most recent cycle
            filter(cn %in% or_cond$plt_cn) %>% #Only single condition plots 
            left_join(or_cond, by = c("cn" = "plt_cn")) %>% 
            filter(fortypcd == 201) 

df_trees <- or_fia[["tree"]] %>% 
            filter(plt_cn %in% df_plots$cn) %>% 
            filter(spcd == 202) %>% 
            filter(!is.na(ht)) %>%
            select(plt_cn, ht, bhage, tpa_unadj, cclcd)

df_site_index <- df_trees %>% 
                 group_by(plt_cn) %>% 
                 summarise(
                     q80 = Hmisc::wtd.quantile(ht, 0.8, weight = tpa_unadj),
                     plot_age = round(mean(bhage, na.rm = TRUE)), #Mean plot age based on cored trees
                     dom_ht_1 = weighted.mean(ht[cclcd %in% c(2, 3)], w = tpa_unadj[cclcd %in% c(2, 3)]), 
                     dom_ht_2 = weighted.mean(ht[ht >= q80], w = tpa_unadj[ht >= q80]),
                     .groups = "drop"
                 ) %>% 
                 gather(value = "dom_ht", key = "method", -plt_cn, -q80, -plot_age) 

or_boundary <-  states() %>% filter(NAME == "Oregon")

ggplot() +
  geom_sf(data = or_boundary, fill = NA) +
  geom_sf(data = df_plots) + 
  theme_bw()

df_site_index %>% 
    mutate(
        group_lab = case_when(method == "dom_ht_1" ~ "FIA CCLCD",
                                        method == "dom_ht_2" ~ "80th quantile mean")
    ) %>%
    ggplot(aes(x = plot_age, y = dom_ht, col = group_lab)) + 
    geom_point() +
    theme_bw() +
    xlab("plot age") +
    ylab("dominant height (feet)")
```

After filtering down only to plots in the Douglas-fir forest type, as well as applying some other basic filters to grab a clean 
pool of FIA data, we end with about 350 plots (Figure 1). A simple scatterplot of the two dominant height estimates vs. age (Figure 2) 
suggests that we can expect some difference in fitted height/age curves. So, we'll go ahead and fit some basic exponential models so we can 
compare results. 
```{r models}
mod_formula <- bf(dom_ht ~ exp(a + b / plot_age),
                  a + b ~ 1,
                  nl = TRUE)

priors <- prior(normal(0, 2), nlpar = "a") +
          prior(normal(0, 2), nlpar = "b") 

model_1 <- brm(mod_formula, prior = priors, data = df_site_index %>% filter(method == "dom_ht_1"))
model_2 <- brm(mod_formula, prior = priors, data = df_site_index %>% filter(method == "dom_ht_2"))

model_1_draws <- df_site_index %>% 
                    filter(method == "dom_ht_1") %>% 
                    add_linpred_draws(model_1) %>% 
                    group_by(plt_cn, plot_age, method, dom_ht) %>%
                    summarise(
                        dom_ht_pred = median(.value),
                        .groups = "drop"
                    ) 

model_2_draws <- df_site_index %>% 
                    filter(method == "dom_ht_2")  %>% 
                    add_linpred_draws(model_2) %>% 
                    group_by(plt_cn, plot_age, method, dom_ht) %>%
                    summarise(
                        dom_ht_pred = median(.value),
                        .groups = "drop"
                    ) 
 


pred_frame <- bind_rows(model_1_draws, model_2_draws) %>% 
              mutate(
                  group_lab = case_when(method == "dom_ht_1" ~ "FIA CCLCD",
                                        method == "dom_ht_2" ~ "80th quantile mean")
              )

glimpse(pred_frame)
```

```{r plot}

pred_frame %>% 
  ggplot(aes(x = plot_age, y = dom_ht))  +
  geom_line(aes(y = dom_ht_pred, color = group_lab)) + 
  theme_bw() +
  theme(legend.title=element_blank()) +
  xlab("plot age (years)") +
  ylab("dominant height (feet)")

```

The results of the fitted models are shown in Figure 3. As you can see, the model fit to dominant height as estimated using the 
80th quantile expects a much higher asymptote (around 15 feet) and generally greater dominant height predictions across much of 
the age range. As you can see, the choice of our observation model (dominant height calculation in this case) will mean a pretty 
big difference in our final estimates. If we were to apply these models in a management context, they'd tell pretty different 
stories about site potential across the state of Oregon. 

Of course, this is an overly simple example to make any real conclusions, but the point is that we ended up with different outcomes 
just by changing the observation model. This is important to highlight, because as practicing biometricians we make decisions about 
observation models all the time. Every time you select a volume table, apply a VBAR, or choose a biomass expansion factor, you're making 
a statement about how the data you're collecting relates to the value you want to estimate. In many cases, you may have a piece of 
software make these decisions for you! 

I'm not saying this is bad; "conventional methods" become so because they're consistent and tend to work well in many contexts. 
But I do think it's worth considering the impact of these choices. Think about the story we started off with. We saw some results 
that didn't pass the smell test, and understanding how the "measurements" (plot estimates of dominant height) were being derived
was critical to landing on the solution. Making sure you understand your observation model, and therefore have a clear idea of how 
your data relate to the attributes you want to estimate, can lead to better inferences and more confident decision-making. 

As always, we welcome feedback: brian@ncx.com 

